Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-05-17 main/MSstatsPTM_Main.tex
Updated main manuscript

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary \todo{link to Sec. S4.1}. 
In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate \todo{figref Fig. 4a}. However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig. 4b)}. As shown in \todo{figref Fig. 4c-e}, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data \todo{figref (Fig. 4d-e)}. In cases with small sample sizes (e.g., 2 replicates) as shown in \todo{figref Fig. 4d}, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in \todo{figref Fig. 4e} [with 4 conditions] over \todo{figref Fig. 4d} [with 2 conditions]). Further results are provided in Supplementary \todo{link to Sec. S4.1}.

@@

PTM experiments often create batches during data acquisition. Whereas the proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects; t-test is not directly applicable to a problem with batches of data. Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch. Although simple, these ad-hoc methods lack statistical justification. We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec. S4.2}. Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch). 
The major results are summarized in \todo{figref Fig. 5} with additional details provided in Supplementary \todo{link to Sec. S4.2}. In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig. 5a). In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig. 5b). As shown Fig. 5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability. Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig. \todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs. S8-S9}). The negative impact was only partially reduced by increasing the sample size to 5. The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.

@@

We evaluated our approach using a custom designed benchmark experiment, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. \todo{Add figure of mixtures} Two sets of data were acquired for each mixture: modifed peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate. These datasets were used as the PTM and global protein data.

@@

In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment they follow the expected fold change much closer. Additionally, background peptides that should not show a significant change, colored grey, show many false positives before adjustment is made. Again after adjustment the results improve, and the number of false positives decrease significantly. In this experiment adjusting to remove confounding with changes in global protein abundance leads to a more accurate conclusion about the effect of modified peptides between conditions.

@@

We evaluated our approach using a custom designed benchmark experiment, where...
Two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared. Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul. The experiment resulted in 15 KGG and 15 global profiling runs. To correct for run-to-run variation, the AQUA peptide mixture was spiked into each sample. There are 4696 ubiquitinated proteins identified in this dataset. Among these, 3173 have corresponding measurements from global data.
To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4. Systematic changes in the abundance of ubiquitinated sites between the groups are present in the KGG data. These changes are considered as artifacts driven by the changes in the abundance of their corresponding proteins. Using the comparison of K2P2 vs. K1P1 as example, there are 19355 possible comparisons with KGG data, and 3860 changes are detected. Out of the 19355 sites, 15036 have corresponding measurements from global data and are eligible for protein-level correction. The correction significantly reduced the number of false positives to 79. Similar observations were made in the other two comparisons K4P4 vs. K2P2 and K4P4 vs. K1P1, as shown in Table XX.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Ubiquitin changes in bone marrow derived macrophages from Atg16L1 deficient mice Re-analysis of published dataset: USP30}

Compare results with/without adjustment\\
Very few measurements of unmodified peptides are available\\
Compare results with original findings] [Need original findings\\
Include USP30OE vs. Ctrl in addition to Combo vs. CCCP?\\
Comment on features, degrees of freedom\\
Different changes across batches – P54652, Q5VYK3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Re-analysis of published dataset: RIP1 [Need this dataset and original findings]}

@@

To evaluate our approach on TMT experiments we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \todo{Add Shigella paper as source} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri. The TMT acquisition was run with two mixtures consisting of 11 samples over an extended time period. Samples were taken at three seperate time periods, for a total of three test conditions. A global profiling run was also conducted, which was used to correct for changes in overall protein abundance.

>>>>

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary \todo{link to Sec. S4.1}.

In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate \todo{figref Fig. 4a}. However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig. 4b)}. As shown in \todo{figref Fig. 4c-e}, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data \todo{figref (Fig. 4d-e)}. In cases with small sample sizes (e.g., 2 replicates) as shown in \todo{figref Fig. 4d}, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in \todo{figref Fig. 4e} [with 4 conditions] over \todo{figref Fig. 4d} [with 2 conditions]). Further results are provided in Supplementary \todo{link to Sec. S4.1}.

@@

PTM experiments often create batches during data acquisition. The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data. Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.\todo{Find source for batch analysis} Although simple, these ad-hoc methods lack statistical justification. We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec. S4.2}. Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).

The major results are summarized in \todo{figref Fig. 5} with additional details provided in Supplementary \todo{link to Sec. S4.2}. In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig. 5a). In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig. 5b). As shown Fig. 5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability. Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig. \todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs. S8-S9}). The negative impact was only partially reduced by increasing the sample size to 5. The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. \todo{Add figure of mixtures} Two sets of data were acquired for each mixture: modifed peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate. These datasets were used as the PTM and global protein data.

@@

In *FigXX* we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is much close to expectation. Additionally, background peptides that should not show a significant change, colored grey, show many false positives before adjustment is made. Again after adjustment the results improve, and the number of false positives decrease significantly. In this experiment adjusting to remove confounding with changes in global protein abundance leads to a more accurate conclusion about the effect of modified peptides between conditions.

@@

We evaluated our approach using a custom designed dilution benchmark experiment where two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared. Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul. The experiment resulted in 15 KGG and 15 global profiling runs. To correct for run-to-run variation, the AQUA peptide mixture was spiked into each sample. There are 4696 ubiquitinated proteins identified in this dataset. Among these, 3173 have corresponding measurements from global data.
To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4. Systematic changes in the abundance of ubiquitinated sites between the groups are present in the KGG data. These changes are considered as artifacts driven by the changes in the abundance of their corresponding proteins. Using the comparison of K2P2 vs. K1P1 as example, there are 19355 possible comparisons with KGG data, and 3860 changes are detected. Out of the 19355 sites, 15036 have corresponding measurements from global data and are eligible for protein-level correction. The correction significantly reduced the number of false positives to 79. Similar observations were made in the other two comparisons K4P4 vs. K2P2 and K4P4 vs. K1P1, as shown in Table XX. \todo{Create or find this table}

@@

To evaluate our approach on TMT experiments we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \cite{Maculins} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri. 11-plex isobaric multiplex was used with two separate mixtures. Cell lysates were prepared over three different time periods, either uninfected, infected at early (45-60 min) or infected at late (3-3.5 hour) time points. There were 103,700 peptides mapping to 9,430 proteins in the global profiling run. Additionally, there were 25,600 unique phosphorylation sites and 12,400 KGG ubiquitination sites. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

The major results are summarized below.
